Forecasting the US Election 2024

### 2024-11-03 paper/paper.qmd
Added paper.pdf and paper.qmd

---
title: "Forecasting the US Election 2024"
subtitle: "Modeling the U.S. Polling Data to forecast the outcome of he 2024 U.S. Presidential Elections"
author: 
  - Maryam Ansari
  - Amy Jin
  - Maggie Zhang

@@

# Introduction

@@

The remainder of this paper is structured as follows. @sec-data....

@@

We use the polling data to forecast the potential outcomes of the 2024 U.S. presidential election between Kamala Harris and Donald Trump.The dataset was obtained from https://projects.fivethirtyeight.com/polls/president-general/2024/national/ or (FiveThirtyEight’s “Poll of Polls” for the 2024 U.S. Presidential election [@fivethirtyeight2024]). It provides a comprehensive view of voter preferences through aggregated results from numerous national polls conducted by various polling organizations. The dataset was simulated, cleaned, analyzed, and tested using the R programming language [@talia], tidyverse [@citetidyverse], knitr [@citeknitr], ggplot2 [@citeggplot2] for plots, gt[@gt] for tables, tidyr[@R-tidyr], arrow[@R-arrow] for parquet, here[@R-here], rstanarm[@R-rstanarm], broom[@R-broom], loo[@R-loo], lubridate [@R-lubridate], while tibble [@R-tibble] helped simplify data frame management. The testthat package [@R-testthat] was essential for unit testing and ensuring code reliability, and we employed styler [@R-styler] for reformatting and maintaining a consistent code style.

@@

**poll_id:** a unique identifier for each poll conducted pollster: which indicates the organization conducting the poll

**sample_size:** representing the total number of respondents

**population:** specifying the voting group described (e.g., likely voters)

**candidate_name:** the names of the candidates in the poll (e.g., Kamala Harris, Donald Trump)

**pct:** the percentage of the vote or support received by each candidate.

These variables allow us to explore various dimensions of polling data, such as trends in voter support across different states and the influence of pollster reliability on polling outcomes. The table below provides a snapshot of the polling data, displaying the first ten entries. This includes the pollster names, sample sizes, and the percentage of support for each candidate, which can reveal patterns in public opinion and help identify how different organizations may report varying levels of support for Kamala Harris and Donald Trump.

TODO: descibe

```{r}
#| label: tbl-pct
#| tbl-cap: "Histogram of Trump Support Percentage (pct)"
#| echo: false
ggplot(data, aes(x = pct)) +
  geom_histogram(binwidth = 5, fill = "#e31a1c", color = "black") +
  theme_minimal() +
  labs(
    x = "Percentage Support (pct)",
    y = "Frequency"
  )
```

**Pollster**: The name of the polling organization that conducted the poll (e.g., YouGov, RMG Research).

**State:** The U.S. state where the poll was conducted.

**Sample Size:** The total number of respondents that participated in the poll. End Date: The date the poll ended.

@@

### Data Manipulation and cleaning:

Filtered out lower quality polls by only keeping polls that met a set numeric grade threshold (numeric_grade \>= 2.5). Filtered out pollsters that had less than 5 entries. Filtered out data that was collected before July 21st, 2024. Added a variable num_trump to count the number of votes for Donald Trump using pct. For data entries where the state was missing, they were categorised as “National”, to standardize the data.

## Summary statistics & Relationships

@@

The relationship between polling organization reliability, as indicated by the pollscore, and the percentage of support (pct) for each candidate was examined. It is anticipated that more reliable pollsters will yield more accurate predictions. Consequently, polls were categorized into three tiers based on their pollscore: high reliability (pollscore greater than 0), medium reliability (pollscore between -1 and 0), and low reliability (pollscore less than -1). By comparing the average support percentages (pct) for Kamala Harris and Donald Trump across these tiers, this analysis sought to identify whether more reliable pollsters produce different outcomes than those deemed less reliable. Mean pct values were calculated for each category, and standard deviation was included to illustrate variability within the poll results. The table (@tbl-reliability) and bar chart @fig-reliability showing the average `pct` for each candidate across different levels of pollster reliability. Discussion: This analysis highlights whether pollsters with higher reliability scores offer more accurate predictions and if their estimates favor one candidate over the other.

```{r}
#| label: tbl-reliability
#| tbl-cap: "Summary Statistics for Support Percentages Across Pollster Reliability Levels"
#| echo: false
# Categorizing poll reliability
# Create a new dataset with categorized reliability
# Remove rows with NA values in the 'pollscore' column while keeping the rest of the data frame
pollscore_data <- data %>%
  filter(!is.na(pollscore)) %>%
  mutate(
    reliability = case_when(
      pollscore > 0 ~ "High",
      pollscore > -1 & pollscore <= 0 ~ "Medium",
      pollscore <= -1 ~ "Low"
    )
  )

@@

```{r}
#| label: fig-reliability
#| fig-cap: "Average Support Percentage by Pollster Reliability"
#| echo: false

# Create a more aesthetic bar chart with alpha and no labels
ggplot(summary_data, aes(x = reliability, y = mean_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7, color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_pct - sd_pct, ymax = mean_pct + sd_pct), 
                width = 0.2, 
                position = position_dodge(0.7)) +
  labs(
    subtitle = "Comparison of support percentages for Kamala Harris and Donald Trump",
    x = "Pollster Reliability",
    y = "Average Support (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) + # Custom colors
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

Polling methodology is a critical factor that influences the results of any survey. Polls were categorized according to their methodology (e.g., Online Panel, Phone Interview) to evaluate how different polling methods might affect the percentage of support (pct) for each candidate. For each polling methodology, the average pct for Kamala Harris and Donald Trump was calculated and compared. This analysis aimed to determine whether specific methodologies consistently resulted in higher or lower support for either candidate. Comparison of candidate vote percentages by polling methodology is demonstrated by [@tbl-methodology] Discussion: We observed potential biases in support based on the polling method, providing insights into which methodologies might offer more reliable forecasts.

```{r}
#| label: tbl-methodology
#| tbl-cap: "Table of Methodologies comparing the Mean Support Percentage for Donald Trump and Kamala Harris with Difference (Kamala Harris - Donald Trump)"
#| echo: false

@@

The impact of sample size (sample_size) on the accuracy of polling results was examined. Larger sample sizes are typically considered more reliable, prompting the categorization of polls into small, medium, and large groups based on sample size. For each group, the average percentage of support (pct) for Kamala Harris and Donald Trump was calculated, and variability within each sample size category was analyzed to evaluate the reliability of the results. A chart[@fig-samplesize] comparing candidate vote percentages by sample size category. Discussion: The results suggest whether larger sample sizes produce more accurate and reliable forecasts, helping us to understand the potential limitations of smaller polls.

```{r}
#| label: fig-samplesize
#| fig-cap: "Candidate Vote Percentages by Sample Size Category"
#| echo: false
# Categorizing sample size into small, medium, and large groups

@@

### Sponsorship and Bias in Polling

Finally, the potential bias of polls sponsored by partisan organizations was assessed to determine whether such sponsorship influenced support for the associated candidate. The vote percentages (pct) for Kamala Harris and Donald Trump were compared in polls sponsored by Democratic-leaning and Republican-leaning organizations (sponsor_candidate_party). By calculating the average pct for each candidate within partisan-sponsored polls, this analysis explored whether partisan sponsorship resulted in a systematic overestimation of support for a particular candidate. A comparison of the polling results in partisan-sponsored polls, broken down by party affiliation is shown by [@fig-sponsorship]. Discussion: This analysis reveals potential biases in partisan-sponsored polls and assesses the objectivity of different polling organizations.

```{r}
#| label: fig-sponsorship
#| fig-cap: "Vote Percentages by Partisan Sponsorship"
#| echo: false
# Calculate the average pct and standard deviation for each candidate by sponsorship type
sponsorship_summary_data <- data %>%
  filter(sponsor_candidate_party != "Non-Partisan") %>%  # Focus on partisan-sponsored polls
  group_by(candidate_name, sponsor_candidate_party) %>%
  summarise(
    avg_pct = mean(pct),
    sd_pct = sd(pct),
    .groups = 'drop'
  )

# Create the bar chart comparing vote percentages by sponsorship type
ggplot(sponsorship_summary_data, aes(x = sponsor_candidate_party, y = avg_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7, color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymin = avg_pct - sd_pct, ymax = avg_pct + sd_pct), 
                width = 0.2, 
                position = position_dodge(0.7)) +
  labs(
    x = "Sponsorship (Party Affiliation)",
    y = "Average Support Percentage (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

# Statistical Model and Analysis

@@

#### Model Specifications

@@

### We mathematically define these models as follows:

@@

#### Variables and Justification

1.  **Dependent Variable**: ( y_i ) (support percentage, `pct`).
2.  **Predictors**:
    -   **(** \text{end\_date} ): Captures time trends, which are essential in understanding shifts in public opinion.
    -   **(** \text{pollster} ): Accounts for differences in polling methodologies and potential biases.
    -   **(** \text{is\_national} ): Differentiates between polls conducted at the national level and those specific to individual states, as national polls tend to exhibit different characteristics and sample compositions compared to state polls.

@@

#### Model Justification

@@

**Model 1** serves as a baseline, capturing the temporal trend.

**Model 2** addresses variability between pollsters, which is crucial given differences in methodologies and sample recruitment.

**Model 3** adds another layer of nuance by accounting for the national vs. state-specific distinction.

**Model 4** employs Bayesian inference, allowing us to incorporate uncertainty and use prior distributions for more robust predictions.

@@

#### Assumptions, Limitations, and Circumstances

@@

#### Software and Model Validation

@@

**AIC/BIC**: For model selection based on goodness-of-fit and penalization for complexity.

**Cross-validation**: (To be detailed further in the appendix) to evaluate out-of-sample performance.

**Residual Diagnostics**: To check for violations of model assumptions.

**Model Convergence**: Monitored using diagnostics provided by `rstanarm` for the Bayesian model.

@@

#### Strengths, Weaknesses, and Final Model Choice

@@

# Results

@@

**2. Time-Based Trend Analysis:** By incorporating poll end dates, our model captures temporal trends, allowing us to observe shifts in candidate support over time. Figure 1 displays the predicted support levels for each candidate over the final months leading up to the election, highlighting any momentum or shifts that may favor one candidate. This visualization provides insight into how public opinion has evolved closer to the election date.

**3. Adjustment for Pollster-Specific Biases:** Pollster biases can significantly influence polling averages. Our model addresses these variations through the pollster-specific coefficients \( \gamma_j \), which adjust each pollster's results based on historical tendencies. Figure 2 illustrates these adjustments, showing how certain pollsters tend to lean in favor of one candidate over another. This correction improves the reliability of our aggregated forecast by balancing these systematic biases.

**4. Estimation of Overall Support:** Aggregating the predicted support from all polls, we computed an average prediction for each candidate as election day nears. This aggregation provides a summary of anticipated national-level support based on polling data and adjusted for timing and pollster effects (see Table 2 for summary statistics of aggregated support levels). This result allows for a more precise comparison of each candidate's standing in the final stretch of the campaign.

**5. Interpretation of Results:** The analysis suggests that [Candidate A] holds a lead in projected support, with a predicted margin of [X percentage points] over [Candidate B]. This result implies a favorable position for [Candidate A], assuming that polling trends continue as projected. While this forecast is based on aggregate polling data, the adjustments for time trends and pollster effects lend robustness to the model's predictions, offering a refined projection of likely support levels for each candidate.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

@@

# Appendix 1 {.unnumbered}

@@

# Appendix B: Idealized Survey Methodology and Survey Design

@@

### Idefitiying Strata:

>>>>

---
title: "Forecasting the US Election 2024"
subtitle: "Modeling the U.S. Polling Data to forecast the outcome of the 2024 U.S. Presidential Elections"
author: 
  - Maryam Ansari
  - Amy Jin
  - Maggie Zhang

@@

\newpage
# Introduction

@@

The rest of this paper is organized as follows. @sec-data covers the data utilized in this analysis, highlighting key variables and sources, particularly focusing on the quality metrics that influence polling accuracy. @sec-model details our modeling approach for each candidate, drawing on insights from recent electoral cycles. @sec-results provides our predictions for the Electoral College based on the model outputs.  @sec-discussion examines the implications of our findings and offers suggestions for future research. Finally, the @sec-appendixa and @sec-appendixb assesses YouGov's polling methodology alongside our proposed survey methodology.

@@

We use the polling data to forecast the potential outcomes of the 2024 U.S. presidential election between Kamala Harris and Donald Trump.The dataset was obtained from FiveThirtyEight’s “Poll of Polls” for the 2024 U.S. Presidential election [@fivethirtyeight2024]. It provides a comprehensive view of voter preferences through aggregated results from numerous national polls conducted by various polling organizations.

The dataset was simulated, cleaned, analyzed, and tested using the R programming language [@talia], tidyverse [@citetidyverse], knitr [@citeknitr], ggplot2 [@citeggplot2] for plots, gt[@gt] for tables, tidyr[@R-tidyr], arrow[@R-arrow] for parquet, here[@R-here], rstanarm[@R-rstanarm], broom[@R-broom], loo[@R-loo], lubridate [@R-lubridate], while tibble [@R-tibble] helped simplify data frame management. The testthat package [@R-testthat] was essential for unit testing and ensuring code reliability, and we employed styler [@R-styler] for reformatting and maintaining a consistent code style.

@@

- **poll_id:** a unique identifier for each poll conducted pollster: which indicates the organization conducting the poll
- **sample_size:** representing the total number of respondents
- **population:** specifying the voting group described (e.g., likely voters)
- **candidate_name:** the names of the candidates in the poll (e.g., Kamala Harris, Donald Trump)
- **pct:** the percentage of the vote or support received by each candidate.

These variables allow us to explore various dimensions of polling data, such as trends in voter support across different states and the influence of pollster reliability on polling outcomes. @tbl-firstten below provides a snapshot of the polling data, displaying the first ten entries. This includes the pollster names, sample sizes, and the percentage of support for each candidate, which can reveal patterns in public opinion and help identify how different organizations may report varying levels of support for Kamala Harris and Donald Trump.

@@

```{r, fig.pos='H'}
#| label: tbl-pct
#| tbl-cap: "Histogram of Trump Support Percentage (pct)"
#| echo: false
ggplot(data, aes(x = pct)) +
  geom_histogram(binwidth = 5, fill = "#e31a1c", color = "black") +
  theme_minimal() +
  labs(
    x = "Percentage Support (pct)",
    y = "Frequency"
  )
```

- **Pollster**: The name of the polling organization that conducted the poll (e.g., YouGov, RMG Research).
- **State:** The U.S. state where the poll was conducted.
- **Sample Size:** The total number of respondents that participated in the poll. End Date: The date the poll ended.

## Summary Statistics & Relationships

@@

The relationship between polling organization reliability, as indicated by the pollscore, and the percentage of support (pct) for each candidate was examined. It is anticipated that more reliable pollsters will yield more accurate predictions. Consequently, polls were categorized into three tiers based on their pollscore: high reliability (pollscore greater than 0), medium reliability (pollscore between -1 and 0), and low reliability (pollscore less than -1). By comparing the average support percentages (pct) for Kamala Harris and Donald Trump across these tiers, this analysis sought to identify whether more reliable pollsters produce different outcomes than those deemed less reliable. Mean pct values were calculated for each category, and standard deviation was included to illustrate variability within the poll results. The table (@tbl-reliability) shows the average `pct` for each candidate across different levels of pollster reliability. However, the analysis shows the converse of our hypothesis to be true.

```{r, fig.pos='H'}
#| label: tbl-reliability
#| tbl-cap: "Summary Statistics for Support Percentages Across Pollster Reliability Levels"
#| echo: false
# Categorizing poll reliability
# Create a new dataset with categorized reliability
# Remove rows with NA values in the 'pollscore' column while keeping the rest of the data frame
pollscore_data <- data %>%
  filter(!is.na(pollscore)) %>%
  mutate(
    reliability = case_when(
      pollscore > 0 ~ "High",
      pollscore > -1 & pollscore <= 0 ~ "Medium",
      pollscore <= -1 ~ "Low"
    )
  )

@@

@@

Polling methodology is a critical factor that influences the results of any survey. Polls were categorized according to their methodology (e.g., Online Panel, Phone Interview) to evaluate how different polling methods might affect the percentage of support (pct) for each candidate. For each polling methodology, the average pct for Kamala Harris and Donald Trump was calculated and compared. This analysis aimed to determine whether specific methodologies consistently resulted in higher or lower support for either candidate. Comparison of candidate vote percentages by polling methodology is demonstrated by [@tbl-methodology]. Evidently Harris has higher support across all platforms, but significantly more for mail-to-web, mail-to-phone and probability panel.

```{r, fig.pos='H'}
#| label: tbl-methodology
#| tbl-cap: "Table of Methodologies comparing the Mean Support Percentage for Donald Trump and Kamala Harris with Difference (Kamala Harris - Donald Trump)"
#| echo: false

@@

The impact of sample size *sample_size* on the accuracy of polling results was examined. Larger sample sizes are typically considered more reliable, prompting the categorization of polls into small, medium, and large groups based on sample size. For each group, the average percentage of support *pct* for Kamala Harris and Donald Trump was calculated, and variability within each sample size category was analyzed to evaluate the reliability of the results. [@fig-samplesize] comparing candidate vote percentages by sample size category. Discussion: The results suggest whether larger sample sizes produce more accurate and reliable forecasts, helping us to understand the potential limitations of smaller polls.

```{r, fig.pos='H'}
#| label: fig-samplesize
#| fig-cap: "Candidate Vote Percentages by Sample Size Category"
#| echo: false
# Categorizing sample size into small, medium, and large groups

@@

## Data Manipulation and Cleaning:

- Filtered out lower quality polls by only keeping polls that met a set numeric grade threshold (*numeric_grade* \>= 2.5).
- Filtered out pollsters that had less than 5 entries.
- Filtered out data that was collected before July 21st, 2024.
- Added a variable *num_trump* to count the number of votes for Donald Trump using pct.
- For data entries where the state was missing, they were categorised as “National”, to standardize the data.
- Added a binary variable *is_national*, where a value of 1 indicates that the vote was national.

@@

# Statistical Model and Analysis {#sec-model}

@@

### Model Specifications

@@

### Model Definitions

We mathematically define these models as follows:

@@

## Variables and Justification

1. **Dependent Variable**: $y_i$ which refers to support percentage, `pct`.
2. **Predictors**:
    -  **end_date:** Captures time trends, which are essential in understanding shifts in public opinion.
    -  **pollster:** Accounts for differences in polling methodologies and potential biases.
    -  **is\_national**: Differentiates between polls conducted at the national level and those specific to individual states, as national polls tend to exhibit different characteristics and sample compositions compared to state polls.

@@

## Model Justification

@@

- **Model 1** serves as a baseline, capturing the temporal trend.
- **Model 2** addresses variability between pollsters, which is crucial given differences in methodologies and sample recruitment.
- **Model 3** adds another layer of nuance by accounting for the national vs. state-specific distinction.
- **Model 4** employs Bayesian inference, allowing us to incorporate uncertainty and use prior distributions for more robust predictions.

@@

## Assumptions, Limitations, and Circumstances

@@

## Software and Model Validation

@@

- **AIC/BIC**: For model selection based on goodness-of-fit and penalization for complexity.
- **Cross-validation**: (To be detailed further in the appendix) to evaluate out-of-sample performance.
- **Residual Diagnostics**: To check for violations of model assumptions.
- **Model Convergence**: Monitored using diagnostics provided by `rstanarm` for the Bayesian model.

@@

## Strengths, Weaknesses, and Final Model Choice

@@

# Results {#sec-results}

@@

 **2. Time-Based Trend Analysis:** By incorporating poll end dates, our model captures temporal trends, allowing us to observe shifts in candidate support over time.

 **3. Adjustment for Pollster-Specific Biases:** Pollster biases can significantly influence polling averages. Our model addresses these variations through the pollster-specific coefficients $\gamma_j$, which adjust each pollster's results based on historical tendencies.

 **4. Estimation of Overall Support:** Aggregating the predicted support from all polls, we computed an average prediction for each candidate as election day nears. This aggregation provides a summary of anticipated national-level support based on polling data and adjusted for timing and pollster effects. This result allows for a more precise comparison of each candidate's standing in the final stretch of the campaign.

 **5. Interpretation of Results:** The analysis suggests that Kamala Harris holds a lead in projected support, with a predicted margin of 2% over Donald Trump. This result implies a favorable position for Harris, assuming that polling trends continue as projected. While this forecast is based on aggregate polling data, the adjustments for time trends and pollster effects lend robustness to the model's predictions, offering a refined projection of likely support levels for each candidate.

@@

# Discussion {#sec-discussion}

@@

# Appendix A: YouGov's Polling Methodology {#sec-appendix}

@@

# Appendix B: Idealized Survey Methodology and Survey Design {#sec-appendixb}

@@

### Identifying Strata:
